Framework description and usage

### 2022-03-03 Manuscript/Manuscript_Methods.Rmd
prepare BMC submission

We built an R [@rstat] based framework to simplify two aspects of systematic literature review:
record acquisition and classification.
The code used to generate the results is available at https://github.com/AD-Papers-Material/BART_SystReviewClassifier, while an updated and ready to use version of the framework is distributed as an R package at https://github.com/bakaburg1/BaySREn.
The framework includes several modules that communicate through intermediate outputs stored in standard formats, making it possible for users to extend the framework or easily integrate it with other tools in their pipeline.
See Supplemental Material S1 for an in-depth description of the framework and how to use it.\
The tasks carried out by the framework are grouped into "sessions", which comprise obtaining scientific citation data (records) using a search query and then labelling them as relevant ("positive" in the rest of the text) or not ("negative") for the topic of interest with the help of a machine learning engine (Fig. 1).
The initial search query should be built using domain knowledge, trying to achieve a high relevant/non-relevant record ratio.\
The framework can then generate a new data-driven query from this labelled set to perform a new session to find records possibly missed by the first query.\

```{r method_diagram, fig.cap = "Figure 1. Framework's visual representation.", echo = FALSE, out.width = "100%"}
knitr::include_graphics("methods_diagram.png", auto_pdf = T)
```

@@

The framework models the relevance of a record based on the following fields in the citation data:
title, abstract, authors, keywords, MESH terms [@lipscomb2000medical].
A range of Natural Language Processing (NLP) techniques [@baeza1999modern; @marshall2015systematic; @ananiadou2006text] are employed to convert the textual information in these fields into features for machine learning through a bag-of-words approach [@marshall2015systematic].
Processing of free text fields (title, abstract) includes: tokenisation (i.e., extracting the terms), removal of common stopwords (i.e., sentence components having no semantic value), part-of-speech filtering (only nouns, adjectives, verbs and untagged terms are retained), and lemmatisation of terms (i.e., reduction to their base grammatical form).
Text processing for authors, keywords and MESH terms identifies logical units (e.g., authors' full names, composite keywords) and extracts them.\
Terms appearing in less than 5% of the labelled documents are removed from negative records.
All terms in the positive set are kept to increase sensitivity at the cost of specificity.\
Some terms tend to co-appear in records (non-consecutive ngrams, nc-ngrams), often carrying a particular meaning when they do co-occur.
To detect nc-ngrams, we generated a word network representation [@rousseau2015graph] with edges occurring between terms with a cosine similarity in terms of document co-occurrence > 0.5.
We extracted the maximal cliques in the network [@eppstein2010listing] representing highly correlated groups of terms;
these groups are added to the dataset as individual features.
Only nc-ngrams comprising a maximum of ten terms are kept.\
A second network is built using a co-occurrence threshold of 0.9.
In this case, the cliques represent terms that always appear together and can therefore be considered redundant (i.e., they do not need to be considered separately).
These terms are merged to increase computational efficiency and reduce overfitting.\
The output is a Document-Term Matrix (DTM), with $N_d$ rows representing the records ($D_i$), $N_t$ terms column for the $t_{field}$ terms (divided by record field) and ${0,1}$ values whether $t_{field} \in D_i$.
We also enriched the DTM with features referencing the number of terms in each field to help the model scale term importance based on the field length.

@@

We used a Bayesian Additive Regression Trees (BART) machine learning "classification model" [@chipman2010bart] [in the implementation of @kapelner2013bartmachine] to predict the probability of a record being relevant, given the information coded into the enriched DTM and the initial training set.
We set up the BART model to use 2,000 MCMC iterations (after 250 burn-in iterations) and 50 trees;
we used a $k$ value of 2 to regularise extreme prediction and let the model use missing fields in the DTM as features [@kapelner2015prediction].
Positive records are oversampled ten times to increase sensitivity [@batista2004study].\
The output is a posterior predictive distribution (PPD) for each record describing the probability of it being relevant (i.e., a positive match).
An ensemble of ten models was fitted to improve prediction stability by averaging the PPD between models [@zhou2021ensemble; @dietterich2000ensemble].\

@@

$$\pi_i = \frac{1}{M}\sum_{j=1}^M Pr(L_i = \text{1}|DTM,m_j)$$

@@

$$
\begin{aligned}
\pi_{i,l} = \{\pi_i : Pr(\pi_i) = 1\%\}\\
\pi_{i,u} = \{\pi_i : Pr(\pi_i) = 99\%\}
\end{aligned}
$$

@@

$$U_\pi=[\max\vec{\pi}_{u}^-, \min\vec{\pi}_{l}^+]$$

@@

$$\pi_{i,l} > \max_{\pi \in U_\pi} \pi$$

@@

$$\pi_{i,u} < \min_{\pi \in U_\pi} \pi$$

All other records are labelled as "uncertain".

@@

- A partition tree [@rpart] is built between the DTM and 800 samples from the PPD;
if a term is present multiple times in the DTM (e.g., both in the title and abstract), it is counted just once, and field term count features are removed.
This step generates a list of rules composed by _AND_/_NOT_ "conditions" made of terms/authors/keywords/MESH tokens, which together identify a group of records.
- For each rule, negative conditions (i.e., _NOT_ statements) are added iteratively, starting from the most specific one, until no conditions are found that would not also remove positive records.
- The extended set of rules is sorted by positive-negative record difference in descending order.
The cumulative number of unique positive records is computed and used to group the rules.
Rules inside each group are ordered by specificity.
- The researcher is then asked to review the rule groups and select one or more rules from each group or edit overly specific rules (e.g., citing a non-relevant concept casually associated with a paper, like a numeric value or indicator).
It is possible to exclude a group of rules altogether, especially those with the poorest sensitivity/specificity ratio.
- The selected rules are joined together by _OR_ statements, defining a subset of records with a sensibly higher proportion of positive records than the original set
- Redundant (i.e., rules whose positive records are already included in more specific ones) and non-relevant rules (i.e., conditions that when removed do not impact sensitivity and specificity) are removed.
- Finally, the rules are re-elaborated in a query that can be used to perform a new citation search.

@@

- Size of the initial training set: 50, 100, 250, 500 records;
- Number of models in the ensemble: 1, 5, 10, 20, 40, 60 repetitions;
- Oversampling rate of positive records: (1x (i.e., no oversampling), 10x, 20x;
- PrI quantiles for building the uncertainty zone: 80%, 90%, 98%;
- Source of randomness between models in the ensemble: MCMC sampling only [@robert2004monte], MCMC plus data bootstrapping [@breiman1996bagging] of the training set.

@@

>>>>

We built an R [@rstats2021] based framework to simplify two aspects of systematic literature review:
record acquisition and classification.
The code used to generate the results is available at https://github.com/AD-Papers-Material/BART_SystReviewClassifier, while an updated and ready to use version of the framework is distributed as an R package at https://github.com/bakaburg1/BaySREn.
The framework includes several modules that communicate through intermediate outputs stored in standard formats, making it possible for users to extend the framework or easily integrate it with other tools in their pipeline.
See Supplemental Material S1 for an in-depth description of the framework and how to use it.\
The tasks carried out by the framework are grouped into "sessions", which comprise obtaining scientific citation data (records) using a search query and then labelling them as relevant ("positive" in the rest of the text) or not ("negative") for the topic of interest with the help of a machine learning engine (Fig. 1).
The initial search query should be built using domain knowledge, trying to achieve a high relevant/non-relevant record ratio.\
The framework can then generate a new data-driven query from this labelled set to perform a new session to find records possibly missed by the first query.\

```{r figure1.method_diagram, fig.cap = "Figure 1. Framework's visual representation.", echo = FALSE, out.width = "100%"}
knitr::include_graphics(file.path("Figures", "methods_diagram.png"), auto_pdf = T)
```

@@

The framework models the relevance of a record based on the following fields in the citation data:
title, abstract, authors, keywords, MeSH terms [@lipscomb2000medical].
A range of Natural Language Processing (NLP) techniques [@baeza1999modern; @marshall2015systematic; @ananiadou2006text] are employed to convert the textual information in these fields into features for machine learning through a bag-of-words approach [@marshall2015systematic].
Processing of free text fields (title, abstract) includes: tokenisation (i.e., extracting the terms), removal of common stopwords (i.e., sentence components having no semantic value), part-of-speech filtering (only nouns, adjectives, verbs and untagged terms are retained), and lemmatisation of terms (i.e., reduction to their base grammatical form).
Text processing for authors, keywords and MeSH terms identifies logical units (e.g., authors' full names, composite keywords) and extracts them.\
Terms appearing in less than 5% of the labelled documents are removed from negative records.
All terms in the positive set are kept to increase sensitivity at the cost of specificity.\
Some terms tend to co-appear in records (non-consecutive ngrams, nc-ngrams), often carrying a particular meaning when they do co-occur.
To detect nc-ngrams, we generated a word network representation [@rousseau2015graph] with edges occurring between terms with a cosine similarity in terms of document co-occurrence > 0.5.
We extracted the maximal cliques in the network [@eppstein2010listing] representing highly correlated groups of terms;
these groups are added to the dataset as individual features.
Only nc-ngrams comprising a maximum of ten terms are kept.\
A second network is built using a co-occurrence threshold of 0.9.
In this case, the cliques represent terms that always appear together and can therefore be considered redundant (i.e., they do not need to be considered separately).
These terms are merged to increase computational efficiency and reduce overfitting.\
The output is a Document-Term Matrix (DTM), with $N_d$ rows representing the records ($D_i$), $N_t$ terms column for the $t_{field}$ terms (divided by record field) and ${0,1}$ values whether $t_{field} \in D_i$.
We also enriched the DTM with features referencing the number of terms in each field to help the model scale term importance based on the field length.

@@

We used a Bayesian Additive Regression Trees (BART) machine learning "classification model" [@chipman2010bart] [in the implementation of @kapelner2013bartmachine] to predict the probability of a record being relevant, given the information coded into the enriched DTM and the initial training set.
We set up the BART model to use 2,000 MCMC iterations (after 250 burn-in iterations) and 50 trees;
we used a $k$ value of 2 to regularise extreme prediction and let the model use missing fields in the DTM as features [@kapelner2015prediction].
Positive records are oversampled ten times to increase sensitivity [@batista2004study].\
The output is the expected value posterior predictive distribution for each record (PPD in brief) describing the probability of it being relevant (i.e., a positive match).
An ensemble of ten models was fitted to improve prediction stability by averaging the PPD between models [@zhou2021ensemble; @dietterich2000ensemble].\

@@

$$\pi_i = \frac{1}{M}\sum_{j=1}^M Pr(L_i = \text{1}|DTM,m_j) \tag{Eq. 1}$$

@@

$$
\begin{aligned}
\pi_{i,l} = \{\pi_i : Pr(\pi_i) = 1\%\}\\
\pi_{i,u} = \{\pi_i : Pr(\pi_i) = 99\%\}
\end{aligned}
\tag{Eq. 2}
$$

@@

$$U_\pi=[\max\vec{\pi}_{u}^-, \min\vec{\pi}_{l}^+] \tag{Eq. 3}$$

@@

$$\pi_{i,l} > \max_{\pi \in U_\pi} \pi \tag{Eq. 4}$$

@@

$$\pi_{i,u} < \min_{\pi \in U_\pi} \pi \tag{Eq. 5}$$

The remaining records are labelled as "uncertain".

@@

- A partition tree [@rpart] is built between the DTM and 800 samples from the PPD;
if a term is present multiple times in the DTM (e.g., both in the title and abstract), it is counted just once, and field term count features are removed.
This step generates a list of rules composed by _AND_/_NOT_ "conditions" made of terms/authors/keywords/MeSH tokens, which together identify a group of records.
- For each rule, negative conditions (i.e., _NOT_ statements) are added iteratively, starting from the most specific one, until no conditions are found that would not also remove positive records.
- The extended set of rules is sorted by positive-negative record difference in descending order.
The cumulative number of unique positive records is computed and used to group the rules.
Rules inside each group are ordered by specificity.
- The researcher is then asked to review the rule groups and select one or more rules from each group or edit overly specific rules (e.g., citing a non-relevant concept casually associated with a paper, like a numeric value or indicator).
It is possible to exclude a group of rules altogether, especially those with the poorest sensitivity/specificity ratio.
- The selected rules are joined together by _OR_ statements, defining a subset of records with a sensibly higher proportion of positive records than the original set
- Redundant (i.e., rules whose positive records are already included in more specific ones) and non-relevant rules (i.e., conditions that when removed do not impact sensitivity and specificity) are removed.
- Finally, the rules are re-elaborated in a query that can be used to perform a new citation search.

@@

- Size of the initial training set: 50, 100, 250, 500 records;
- Number of models in the ensemble: 1, 5, 10, 20, 40, 60 repetitions;
- Oversampling rate of positive records: 1x (i.e., no oversampling), 10x, 20x;
- PrI quantiles for building the uncertainty zone: 80%, 90%, 98%;
- Source of randomness between models in the ensemble: MCMC sampling only [@robert2004monte], MCMC plus data bootstrapping [@breiman1996bagging] of the training set.

@@

## Software and data
The framework is built with R v4.0.4 [@rstats2021]. The R packages required by the framework are listed at https://github.com/bakaburg1/BaySREn/blob/main/DESCRIPTION. All relevant data necessary to replicate the results is available at https://doi.org/10.5281/zenodo.6323360.

### 2022-03-03 Manuscript/Manuscript_Results.Rmd
prepare BMC submission

knitr::opts_chunk$set(
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 400,
  cache = TRUE,
  cache.rebuild = FALSE
)
```

@@

*`r first_query`*

@@

We extracted the 15 more relevant terms for the classification model, described as:
Term (citation part): Inclusion Rate (Inclusion Stability) [linear Relative Risk, Statistic].

`r var_importance`.

@@

*`r new_query`*

@@

```{r performance_plot, cache = FALSE, fig.width=7.5, fig.height=4.5, fig.cap="**Figure 2**. Observed cumulative number of positive matches (red dots) sorted by simple query ordering. The [trunc. 90% PrI] of the cumulative positive matches estimated by the Bayesian logistic model is shown as a shaded area delimited by the 95% quantile of the PrI and by the observed number of positive matches (light blue lines). A darker blue line represents the median of the PrI."}

>>>>

knitr::opts_chunk$set(
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 400,
  cache = TRUE,
  fig.path = "Figures/",
  cache.rebuild = FALSE
)
```

@@

*`r first_query`*\

@@

We extracted the 15 more relevant terms for the classification model, described as:
Term (citation part): Inclusion Rate (Inclusion Stability) [linear Relative Risk, Statistic]:\

`r var_importance`.\

@@

*`r new_query`*\

@@

```{r figure2.performance_plot, cache = FALSE, fig.width=7.5, fig.height=4.5, fig.cap="**Figure 2**. Observed cumulative number of positive matches (red dots) sorted by simple query ordering. The [trunc. 90% PrI] of the cumulative positive matches estimated by the Bayesian logistic model is shown as a shaded area delimited by the 95% quantile of the PrI and by the observed number of positive matches (light blue lines). A darker blue line represents the median of the PrI."}

### 2022-03-03 Manuscript/S1.Framework_description.Rmd
prepare BMC submission

---
title: "Study protocol"
date: "`r lubridate::today()`"
output:
  pdf_document:
    keep_tex: true
    toc: yes
  github_document:
    toc: yes
---

@@

# Research Protocol

@@

To use the framework clone it from https://github.com/AD-Papers-Material/BART_SystReviewClassifier and run the R project in the repository.

@@

- The framework extracts **features** from text data (i.e., title and abstract) of the records, authors, keywords and MESH terms (see Methods for more info).
Before features extraction, text data terms are lemmatised, stop-words are removed, and only nouns, adjectives, verbs and untagged terms are retained.
Authors, keywords, and MESH are kept unmodified.
Co-occurrent features (non-consecutive ngrams, **nc-grams**) are identified and **redundant features** (i.e. terms always appearing together) are joined to decrease the feature space noise.
- For each term, a 0/1 label specifies if it is present in each record.
- Positive records are oversampled **10x** to improve sensitivity;
- An extra feature is added for each record field (title, abstract, authors, keywords, MESH terms), reporting the number of terms in each.

@@

- We fit a partition tree between the DTM and 800 samples from the PPD;
if a term is present multiple times in the DTM (e.g., both title and abstract), they are counted just one, and field term count features are removed.
This step generates a list of rules composed by $AND$/$NOT$ "conditions" made of terms/authors/keywords/MESH tokens, which together identify a group of records.
- For each rule, negative conditions (i.e., $NOT$ statements) are added iteratively, starting from the most specific one, until no conditions are found that would not also remove positive records.
- The extended set of rules is sorted by positive-negative record difference in descending order.
The cumulative number of unique positive records is computed and used to group the rules.
Rules inside each group are ordered by specificity.
- The researcher is then asked to review the rule groups, selecting one or more rules (useful if they convey different meaning) from each, or edit them (in case too specific positive or negative conditions were included).
It is possible to exclude a group of rules altogether, especially those with the worse sensitivity/specificity ratio.
- The selected rules are joined together by $OR$ statements, defining a subset of records with a sensibly higher proportion of positive records than the original one.
- Redundant rules (i.e., rules whose positive records are already included in more specific ones) and conditions (i.e., conditions that once removed do not decrease the total number of positive or do not increase the negative records) are removed.
- Finally, the rules are re-elaborated in a query usable on the major scientific databases.

>>>>

---
title: "Framework description and usage"
date: "`r lubridate::today()`"
output:
  pdf_document:
    keep_tex: true
    toc: yes
  github_document:
    toc: yes
---

@@

# Framework usage instructions

@@

To use the framework, clone it from https://github.com/AD-Papers-Material/BART_SystReviewClassifier and run the R project in the repository.

@@

- The framework extracts **features** from text data (i.e., title and abstract) of the records, authors, keywords and MeSH terms (see Methods for more info).
Before features extraction, text data terms are lemmatised, stop-words are removed, and only nouns, adjectives, verbs and untagged terms are retained.
Authors, keywords, and MeSH are kept unmodified.
Co-occurrent features (non-consecutive ngrams, **nc-grams**) are identified and **redundant features** (i.e. terms always appearing together) are joined to decrease the feature space noise.
- For each term, a 0/1 label specifies if it is present in each record.
- Positive records are oversampled **10x** to improve sensitivity;
- An extra feature is added for each record field (title, abstract, authors, keywords, MeSH terms), reporting the number of terms in each.

@@

- We fit a partition tree between the DTM and 800 samples from the PPD;
if a term is present multiple times in the DTM (e.g., both title and abstract), they are counted just one, and field term count features are removed.
This step generates a list of rules composed by $AND$/$NOT$ "conditions" made of terms/authors/keywords/MeSH tokens, which together identify a group of records.
- For each rule, negative conditions (i.e., $NOT$ statements) are added iteratively, starting from the most specific one, until no conditions are found that would not also remove positive records.
- The extended set of rules is sorted by positive-negative record difference in descending order.
The cumulative number of unique positive records is computed and used to group the rules.
Rules inside each group are ordered by specificity.
- The researcher is then asked to review the rule groups, selecting one or more rules (useful if they convey different meaning) from each, or edit them (in case too specific positive or negative conditions were included).
It is possible to exclude a group of rules altogether, especially those with the worse sensitivity/specificity ratio.
- The selected rules are joined together by $OR$ statements, defining a subset of records with a sensibly higher proportion of positive records than the original one.
- Redundant rules (i.e., rules whose positive records are already included in more specific ones) and conditions (i.e., conditions that once removed do not decrease the total number of positive or do not increase the negative records) are removed.
- Finally, the rules are re-elaborated in a query usable on the major scientific databases.
